``Interestingness'' in Mathematical Theory Exploration

### 2019-10-04 PhDThesis/background_theoryexploration.tex
Fold feedback into tex

(Automated) Theory Exploration (TE) is the task of taking a \emph{signature}
of definitions in some formal system (for example a programming language) and
automatically generating a set of formal statements (properties) involving
those definitions. These may be conjectures or theorems (proven either by
sending conjectures to an automated theorem prover, or by having the generating
procedure proceed in logically sound steps). These statements must also be
``interesting'' in some way, to rule out unhelpful trivialities such as
iterating a pattern over and over (e.g. \hs{x + 0 = x}, \hs{(x + 0) + 0 = x},
\dots).

@@

The choice of appropriate methods can differ according to the underlying logic,
etc. that is in use, so in this work we ground ourselves by applying TE to
discovering properties of pure functional software libraries, mostly in the
Haskell programming language. This also gives our results immediate utility to
software engineering. We ignore the issue of \emph{proving} such conjectures,
deferring that task to existing tools (such as \hipspec{}).

The method of conjecture generation is a key characteristic of any theory
exploration system, although all existing implementations rely on brute force
enumeration to some degree. We focus on \quickspec{}~\cite{QuickSpec}, which
conjectures equations about Haskell code. We have used version 1 of
\quickspec{} in our experiments, due to its availability, stability and tooling
integration; hence references to \quickspec{} are to version 1 unless stated
otherwise. At the time of writing there is a \quickspec{} version 2 available,
which has a much improved generation procedure which can be significantly faster
than its predecessor. Whilst \quickspec{} 2 has advanced the state of the art in
theory exploration, our preliminary experience has found that it still suffers
the scaling issues we identify in this work\footnote{We encountered memory
  exhaustion when trying to run some of the examples included with the
  \quickspec{} 2 source code; let alone auto-generated signatures.} Since
\quickspec{} 2 has less integration with the tooling we have used, we leave a
more thorough analysis of it (and related systems such as \speculate{}) for
future work, pending the necessary infrastructure changes that would require.

>>>>

(Automated) Theory Exploration (TE) is the task of taking a \emph{signature}
of definitions \iffalse TODO: Alison Some tools generate their own definitions
\fi in some formal system (for example a programming language) and
automatically generating a set of formal statements (properties) involving
those definitions. These may be conjectures or theorems (proven either by
sending conjectures to an automated theorem prover, or by having the generating
procedure proceed in logically sound steps from a set of axioms). These
statements should also be
``interesting'' in some way, to rule out unhelpful trivialities such as
iterating a pattern over and over (e.g. \hs{x + 0 = x}, \hs{(x + 0) + 0 = x},
\dots).

@@

The choice of appropriate methods can differ according to the underlying logic,
etc. that is in use: in this work we ground ourselves by applying TE to
discovering properties of pure functional software libraries, mostly in the
Haskell programming language. This also gives our results immediate utility to
software engineering. We focus on the generation of interesting conjectures; to
\emph{prove} such conjectures, we defer to existing tools (such as \hip{} and
\hipspec{}).

The method of conjecture generation is a key characteristic of any theory
exploration system, although all existing implementations rely on brute force
enumeration to some degree. We focus on \quickspec{}~\cite{QuickSpec}, which
conjectures equations about Haskell code. We have used version 1 of \quickspec{}
in our experiments, due to its availability, stability and tooling integration;
hence references to \quickspec{} are to version 1 unless stated otherwise. At
the time of writing \iffalse TODO: Alison: Include date \fi there is a
\quickspec{} version 2 available, which has a much improved generation procedure
which can be significantly faster than its predecessor. Whilst \quickspec{} 2
has advanced the state of the art in theory exploration, our preliminary
experience has found that it still suffers the scaling issues we identify in
this work\footnote{Not only did we experience failures with our own experiments,
  we also suffered memory exhaustion when trying to run some of the examples
  included with the \quickspec{} 2 source code.} Since \quickspec{} 2 has less
integration with the tooling we have used, we leave a more thorough analysis of
it (and related systems such as \speculate{}) for future work, pending the
necessary infrastructure changes that would require.
